---
title: "Riffle-ito Cambridge Deployment"
author: "Jeff Walker & Don Blair"
date: "August 1, 2014"
output:
  html_document:
    keep_md: yes
    toc: yes
---

This document summarizes data collected by two [Riffle-ito Water Quality Data Loggers](https://github.com/p-v-o-s/riffle-ito) deployed at a USGS streamflow gage in .

**Purpose**: To determine the stability and battery-lifetime of a single riffle-ito deployment, and to compare the on-board RTC temperature measurements to the DHT22 measurements.

**Description**: The riffle-ito was configured with the sketch above and set to record readings from a DHT22 temperature/humidity sensor every 60 seconds. Three fresh Duracell AA batteries were used to power the riffle-ito. Data were retrieved every few days resulting in short gaps and multiple data files, however the batteries were not changed.  

**Location**: 

## Set Up

First we'll load the R packages used for this summary.

```{r load packages, message=FALSE, warning=FALSE}
library(lubridate)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())
```

## Load Data

The raw data are stored in the `./data` directory. There are multiple files for this 
deployment because I temporarily removed the SD card from the riffle-ito to download
the data. As shown in the [sketch](), whenever the riffle-ito is powered on, it creates a new logging file by incrementing the last two digits by one.

```{r load data}
riffle1 <- read.csv('./data/riffle1Deploy.CSV', as.is=TRUE)
riffle2 <- read.csv('./data/riffle2Deploy.CSV', as.is=TRUE)
head(riffle1)
head(riffle2)
```

Next, we want to parse the datetimes using `lubridate::ymd_hms()` to `POSIXct` objects.

```{r parse dates}
riffle1 <- mutate(riffle1,
                  DATETIME=ymd_hms(DATETIME),
                  RIFFLE='Riffle-1',
                  BATTERY_LEVEL=BATTERY_LEVEL*100)
riffle2 <- mutate(riffle2,
                  DATETIME_REG=ymd_hms(DATETIME_REG),
                  RIFFLE='Riffle-2')
summary(riffle1)
summary(riffle2)
```

The data are currently in a wide format, where each column represents a single variable (see [Tidy Data](http://vita.had.co.nz/papers/tidy-data.pdf) and [Reshaping Data with the reshape Package](http://www.jstatsoft.org/v21/i12/paper) by Hadley Wickham for more information about long/wide formats, and note that `tidyr` is a relatively new package that provides much of the same functionality as the `reshape2` package). 

For plotting, it will be easier to convert to a long format. This can easily be done using the `tidyr::gather` function.

```{r gather}
riffle1 <- gather(riffle1, VAR, VALUE, RTC_TEMP_C:BATTERY_LEVEL) %>%
  filter(DATETIME >= ymd_hm("2014-07-25 12:00"),
         DATETIME <= ymd_hm("2014-08-01 10:30"))
riffle2 <- mutate(riffle2, DATETIME=DATETIME_REG) %>%
  dplyr::select(-DATETIME_UNIX, -DATETIME_REG) %>%
  gather(VAR, VALUE, RTC_TEMP_C, TEMP_C, 
         CONDUCT_FREQ_HERTZ, PHOTORESIST, BATTERY_LEVEL) %>%
  filter(DATETIME >= ymd_hm("2014-07-28 17:30"),
         DATETIME <= ymd_hm("2014-08-01 10:30"))
df <- rbind(riffle1, riffle2) %>%
  mutate(RIFFLE=factor(RIFFLE))
summary(df)
```

The data are now in long format with each row corresponding to one measurement for a single variable.

## Visualizations

We can plot the data with each panel showing one of the variables. The data are colored by the corresponding file/deployment. 

```{r plot_timeseries}
ggplot(df, aes(DATETIME, VALUE, color=RIFFLE)) +
  geom_line() +
  facet_wrap(~VAR, scales='free_y') +
  labs(x='Datetime', y='Value') +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
        legend.position='top')
```

Both Riffle-itos were extracted from the field around 2014-08-01 10:30 AM so we'll drop all data after that. 

We can compare the RTC on-board temperature to the DHT22 temperature for verification. The red line in this figure is a 1:1 line of equality; the blue line is a linear regression. This figure shows that the DHT22 temperature (`TEMP_C`) tends to be about 0.5 degC greater than the RTC temperature (`RTC_TEMP_C`).

```{r plot_compare_temp}
spread(df, VAR, VALUE) %>%
  ggplot(aes(RTC_TEMP_C, TEMP_C, color=RIFFLE)) +
  geom_point() +
  geom_abline(color='red', linetype=2) +
  geom_smooth(method='lm')
```

THe differences between the RTC and DHT22 temperature show an interesting (i.e. non-regular) pattern over time.

```{r plot_temp_diff}
spread(df, VAR, VALUE) %>%
  ggplot(aes(DATETIME, RTC_TEMP_C-TEMP_C, color=RIFFLE)) +
  geom_point()
```

As another comparison, we can plot timeseries of the RTC temperature and the DHT22 temperature on the same figure.

```{r plot_temp_ts}
filter(df, VAR %in% c("RTC_TEMP_C", "TEMP_C")) %>%
ggplot(aes(DATETIME, VALUE, color=VAR)) +
  geom_line() +
  facet_wrap(~RIFFLE, ncol=1)
```

## Light Sensor

```{r plot_light}
filter(df, VAR=='PHOTORESIST') %>%
  ggplot(aes(DATETIME, VALUE)) +
  geom_line() +
  labs(x='Datetime', y='Photoresistor')
```

```{r plot_light_hour}
filter(df, VAR=='PHOTORESIST') %>%
  mutate(DATE=factor(floor_date(DATETIME, unit='day')),
         DECIMAL_TIME=hour(DATETIME)+minute(DATETIME)/60) %>%
  ggplot(aes(DECIMAL_TIME, VALUE, color=DATE)) +
  geom_line() +
  labs(x='Time of Day', y='Photoresistor')
```

## Conductivity Sensor

```{r plot_cond}
filter(df, VAR=='CONDUCT_FREQ_HERTZ') %>%
  ggplot(aes(DATETIME, VALUE)) +
  geom_line() +
  labs(x='Datetime', y='Conductivity Frequency')
```

# Comparison to USGS Gage

Station: [01104455 STONY BROOK, UNNAMED TRIBUTARY 1, NEAR WALTHAM, MA](http://waterdata.usgs.gov/ma/nwis/uv/?site_no=01104455)

[Download File](http://waterdata.usgs.gov/ma/nwis/uv?cb_00060=on&cb_00010=on&cb_00095=on&cb_63680=on&cb_99404=on&format=rdb&site_no=01104455&period=&begin_date=2014-07-25&end_date=2014-08-01)

```{r load usgs}
usgs <- read.table('./data/01104455_20140725.txt', sep='\t', skip=30)
names(usgs) <- c('AGENCY', 'STATION_ID', 'DATETIME', 'TIMEZONE', 'FLOW', 'FLOW_FLAG', 'TEMP_C', 'TEMP_FLAG', 'SPCOND', 'SPCOND_FLAG', 'TURB', 'TURB_FLAG', 'CHLORIDE', 'CHLORIDE_FLAG')
usgs <- select(usgs, DATETIME, FLOW, TEMP_C, SPCOND, TURB, CHLORIDE) %>%
  mutate(DATETIME=ymd_hm(DATETIME))
head(usgs)
```

Convert to long format.

```{r convert usgs long}
usgs <- gather(usgs, VAR, VALUE, FLOW:CHLORIDE)
```

Plot USGS timeseries

```{r plot_usgs_ts}
ggplot(usgs, aes(DATETIME, VALUE)) +
  geom_line() +
  facet_wrap(~VAR, scales='free_y')
```

Compare Riffle temp to USGS temp

```{r plot_usgs_riffle_temp}
riffle.temp <- filter(df, VAR=="TEMP_C") %>%
  mutate(SOURCE=RIFFLE) %>%
  select(DATETIME, SOURCE, VAR, VALUE)
usgs.temp <- filter(usgs, VAR=="TEMP_C") %>%
  mutate(SOURCE='USGS') %>%
  select(DATETIME, SOURCE, VAR, VALUE)
temp <- rbind(riffle.temp, usgs.temp)
ggplot(temp, aes(DATETIME, VALUE, color=SOURCE)) +
  geom_line()
```


# Conclusions

