---
title: "Riffle-ito & Thermistor Submerged Deployment"
author: "Jeffrey D Walker"
date: "August 6, 2014"
output:
  html_document:
    keep_md: yes
    toc: yes
---

This document summarizes data collected using a [Riffle-ito Water Quality Data Logger](https://github.com/p-v-o-s/riffle-ito) connected to a [10K Precision Epoxy Thermistor](http://www.adafruit.com/products/372) and submerged in a jar of water.

**Sketch**: [riffleito_thermistor_logger](https://github.com/walkerjeffd/riffle-ito-apps/blob/d137e07d5ac23a683431f4cc5cf6c404482cfeae/ino/riffleito_thermister_logger/riffleito_thermister_logger.ino)

**Purpose**: To test the ability to measure temperature when submerged. And to make sure the enclosure does not leak.

**Description**: The riffle-ito was configured with the sketch above and set to record readings from a 10K thermistor every 60 seconds. The jar was submerged in hot tap water around noon to induce a change in temperature.

**Location**: The riffle-ito was placed in a large glass jar in kitchen in Brunswick, ME. 

## Set Up

First we'll load the R packages used for this summary.

```{r load packages, message=FALSE, warning=FALSE}
library(lubridate)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())
```

## Load Data

The raw data are stored in the `./data` directory. To load the data, first retrieve a list of the filenames.

```{r get filenames}
filenames <- dir(path='./data', pattern='*.CSV', full.names = TRUE)
filenames
```

The `filenames` vector shows that there is/are `r length(filenames)` file(s).

We can then use the `dplyr::rbind_all` file to automatically loop through this set of filenames, load each file, append a column named `FILE` that stores the filename for each dataset, and finally merge the datasets for each file in a single data frame named `df`.

```{r load files}
df <- rbind_all(lapply(filenames, function (filename) {
  read.csv(filename, as.is=TRUE) %>%
    mutate(FILE=filename)
}))
head(df)
```

Next, we want to parse the datetimes using `lubridate::ymd_hms()` to `POSIXct` objects, and convert the `FILE` column to a factor.

```{r parse dates}
df <- mutate(df,
             DATETIME=ymd_hms(DATETIME),
             FILE=factor(FILE))
summary(df)
```

The data are currently in a wide format, where each column represents a single variable (see [Tidy Data](http://vita.had.co.nz/papers/tidy-data.pdf) and [Reshaping Data with the reshape Package](http://www.jstatsoft.org/v21/i12/paper) by Hadley Wickham for more information about long/wide formats, and note that `tidyr` is a relatively new package that provides much of the same functionality as the `reshape2` package). 

For plotting, it will be easier to convert to a long format. This can easily be done using the `tidyr::gather` function.

```{r gather}
df <- gather(df, VAR, VALUE, RTC_TEMP_C:BATTERY_LEVEL)
head(df)
summary(df)
```

The data are now in long format with each row corresponding to one measurement for a single variable.

## Visualizations

We can plot the data with each panel showing one of the four variables. The data are colored by the corresponding filename. 

```{r plot_timeseries}
ggplot(df, aes(DATETIME, VALUE, color=FILE)) +
  geom_line() +
  facet_wrap(~VAR, scales='free_y') +
  labs(x='Datetime', y='Value') +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
        legend.position='top')
```

The spike in temperature was caused by placing the jar in a hot water bath in the sink at around 12:00. It was then removed from the hot water around 15:00 and placed on the counter to cool.

We can compare the RTC on-board temperature to the thermistor temperature for verification. The red line in this figure is a 1:1 line of equality; the blue line is a linear regression. This figure shows good agreement between the thermistor temperature (`TEMP_C`) and the RTC temperature (`RTC_TEMP_C`).

```{r plot_compare_temp}
spread(df, VAR, VALUE) %>%
  ggplot(aes(RTC_TEMP_C, TEMP_C)) +
  geom_point() +
  geom_abline(color='red', linetype=2) +
  geom_smooth(method='lm')
```

The differences between the RTC and thermistor temperature show an interesting (i.e. non-regular) pattern over time.

```{r plot_temp_diff}
spread(df, VAR, VALUE) %>%
  ggplot(aes(DATETIME, RTC_TEMP_C-TEMP_C)) +
  geom_point()
```

As another comparison, we can plot timeseries of the RTC temperature and the thermistor temperature on the same figure.

```{r plot_temp_ts}
filter(df, VAR %in% c("RTC_TEMP_C", "TEMP_C")) %>%
ggplot(aes(DATETIME, VALUE, color=VAR)) +
  geom_line()
```

The differences between the RTC and thermistor temperatures are probably caused by the RTC being within the enclosure, and the thermistor exposed to the water outside the enclosure.

# Conclusions

Based on these plots, I conclude:

- The VOSS water bottle enclosure did not leak.
- The thermistor was able to measure changes in water temperature and collected stable measurements.
